---
title: "Summarizing your data lab 1"
author: "Data Science 101 Team"
output:
     html_document:
         keep_md: true
---

```{r setup, include=FALSE}
#source('http://datascience101.stanford.edu/profile.R')
knitr::opts_chunk$set(echo = TRUE,cache = TRUE)
```

# Lab 1: Robustness of summaries

## Question 1: how many numbers do you need to change in order to change the mean?

Let us start with 20 values generated from uniform distribution.

```{r}
sample <- runif(n = 20, min = 2, max = 5)
sample
sample_mean <- mean(sample)
sample_mean
```
How many numbers among the 20 values do we need to change if we want the new sample mean to be 4?

How about the following example? How many numbers among the 30 values do we need to change if we want the new sample mean to be 2?

```{r}
sample1 <- rnorm(n = 10, mean = 0, sd = 1)
sample1
sample1_mean <- mean(sample1)
sample1_mean
```
So what's your answer to Question 1?

## Question 2: how many numbers do we need to change in order to change the median?

We will look at the same data.
```{r}
sample_median <- median(sample)
sample_median
ordered_sample <- sort(sample)
ordered_sample
```
- How is this sample_median calculated?
- If we change just one number in this sequence to be 3.5, how many possibiities are there for value of the new median?
- If we change the largest value to be 10, does the median change? If we change the smallest value to be 0, does the median change?
- In general, can we get an arbitrary value of median by changing just one value in this sequence?
- How many numbers do we need to change/add if we want a specific median? For example, please answer this question for scenarios: new_median = 2.072841, 3.127627, 4.075274, 5.


## Question 3: which one is more robust, mean or median?

## Trimmed mean

```{r}
mean(x = sample, trim = 0)
mean(x = sample, trim = 0.1)
mean(x = sample, trim = 0.2)
```

