---
title: "A First Experience with R and Rstudio"
author: "Data Science Team"
output:
  html_document:
    keep_md: yes

---

### R and R studio

- R is a language for data analysis and computing

- R studio is an "enviroment" that makes the use of R easier

- It can be confusing at first because you need to get familiar both with R and the new environment, but overall it actually should make your life easier 

### Prompt line execution vs Rmd files

- R is an **interactive** language: you do not need to run a script, submit it and wait for answer

- The best way to familiarize oneself with this is to use the **Console** in Rstudio

- Let's start creating a variable

```{r}
myvariable<-c(1:10)
```

- And plot it
```{r}
plot (myvariable)
```

- Note that it is useful to use the **zoom** button 

- To understand which commands to use, and to develop an analysis approach it is often most useful to work on the console.

- On the other hand, it is useful to be able to keep track of commands, so that we could re-run them another time

- And to annotate what one is trying to do, so that the receord is meaningful for everyone.

- This is where the Rmd files come in 

- You can find documentation on Rmd [here](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)

- Let's also open the file MyRMarkDown.Rmd

- This file is written in .Rmd

- We will provide .Rmd files for all labs

- In class, however, we will often use Console, to do interactive work.

- Our lectures are also written in .Rmd. When lectures contain significant ammounts of R commands that are not echoed, we will also make available those markdown files

### Outbreak data

We are using these to compare the native R commands and the commands that are redefined in a number of packages developed by the same group that developed Rstudio.

- The native R commands have more intuitive names

- Their syntax is simpler

- The "Rstudio" commands make it easier to carry out a number of tasks that one encounter in data analysis, so that they are often preferred in "data science" environments

- Navigating between the two type of commands can be confusing, but at least at the beginning it is important to orient oneself.

#### We start with the native R commands.

```{r}
snow1<-read.csv("./data/snow.csv")
snow1$Date<-as.Date(snow1$Date)
plot(snow1$Date,snow1$Case,ylab="Number of fatal attacks",xlab="Date")
abline(v=as.Date("1854-09-07"))
```

#### "Rstudio commands"

The "Rstudio commands" are what you can get from the functionality of Rstudio. For example, using the "Import Dataset" button, you can see the following commands.

```{r}
library(readr)
library(ggplot2)
snow1<-read_csv("./data/snow.csv",col_types = cols(Date = col_date(format = "%Y-%m-%d")))
qplot(x=snow1$Date,y=snow1$Case,ylab="Number of fatal attacks",xlab="Date") 
```



### Cell Phone Data

Here we will read in the data from the cell phone/car collision study of
Redelmeier and Tibshirani (1997) , and do some simple analyses in R.
For confidentiality,  some of the data have been randomly perturbed.
However the first 4 variables are untouched.


#### Let's read in the data.
First we see what happens with the "native" R functions

```{r}
celldata = read.csv("./data/celldataNew.csv")
names(celldata)
summary(celldata)
```

We can see that 

- the names get modified slighlty, to avoid using a reserved character "-", 

- different variables are interpreted as numerical or factors


Now let's try with the "Rstudio-type" commands

```{r}
celldata = read_csv("./data/celldataNew.csv")
names(celldata)
summary(celldata)
```
We can see that

- Names are left as such

- There is an explicit declaration of what type each of the variables are

- The default interpretation has much more character than factors, so that we would do better modifying it.


We can compute the dimensions of data table

```{r}
dim(celldata)
```
Note that when the data is read in, it is converted to a type of object called a dataframe.
This is a convenient block form for data, that allows us to index the rows and columns
easily
```{r}
is.data.frame(celldata)
```
Other important kinds of objects include vectors, matrices and lists.
In "Rstudio commands", we actually have created a "tibble," which is just mostly just like a data frame. Here are some other useful things to do when you load a new dataset, to understand what type of data you have.

Print out first few lines of data

```{r}
celldata[1:2, ]

```

Look at the first variable
```{r}
table(celldata$"dayof-on")
```
#### Compute histogram of number of years of driving
```{r}
qplot(celldata$"num-yrs-driving", geom = "histogram", color = I("red"), bins = 15)
```
_Note_: The color `I("red")` specification shows the bar boundaries,
using mere `color = "red"` will also work, but provide a legend that
we don't need. This is a detail relating to how `ggplot2` works.

#### Make boxplots for  number of years of driving by gender
Note that celldata[,"num-yrs-driving"] and celldata$"num-yrs-driving" are almost equivalent. The first will return a data frame with one variable, while the second a vector of values.

```{r}
qplot(x = celldata$gender, y = celldata$"num-yrs-driving", geom = "boxplot",xlab="Gender",ylab="Years of driving")
```





#### Main result, without accounting for "intermittent driving"



  Results| Day before/On Phone | Day before/Not on phone
--- | --- | ---
Day of collision: On phone   | 13 | 157
Day of collision: not on phone  | 24 | 505



```{r}
table(celldata$"daybef-on", celldata$"dayof-on")
```

Now let's define a function to compute the relative risk
```{r}
relRisk = function(celldata) {
    n1 = sum(celldata[, "dayof-on"] == 0  & celldata[,"daybef-on"] == 1)
    n2 = sum(celldata[, "dayof-on"] == 1  & celldata[,"daybef-on"] == 0)
    out = n2 / n1
    return(out)
}
```
And let's try it out on different portions of the dataset. We start with the entire data:

```{r}
relRisk(celldata)
```

Then restrict to different speeds
```{r}
hi = celldata[, "Location"] == "highspeed"
relRisk(celldata[hi, ])
```
```{r}
low = celldata[, "Location"] == "lowspeed"
relRisk(celldata[low, ])
```




### 100m Sprint Data

Here we will analyze data on men's and women's world records for the 100m sprint.


#### Let's read in the data on men

```{r}
library(readr)
data.men = read_delim("./data/100men", 
    "\t", escape_double = FALSE, col_types = cols(Date = col_date(format = "%Y-%m-%d")), 
    trim_ws = TRUE)
data.men
```
#### Let's read in the data on women

```{r}

data.women<- read_delim("./data/100women", 
    "\t", escape_double = FALSE, col_types = cols(Date = col_date(format = "%B %d, %Y")), 
    trim_ws = TRUE)
```

#### Extract numerical times and years

```{r}
time.men = data.men$Time
date.men = data.men$Date
time.men
year.men =as.numeric(format(date.men,'%Y'))

time.women = data.women$Time
date.women = data.women$Date
year.women = as.numeric(format(date.women,'%Y'))
time.women
year.women

```

Reverse order of men's data, to be most recent last (as in women)
```{r}
year.men = rev(year.men)
time.men = rev(time.men)
```

#### Plot data
```{r}
library(ggplot2)
qplot(x = year.men, y = time.men,
      geom = c("point", "line"),
      xlab = "Year", ylab = "Time",
      main = "100m times for men")

```

```{r}
qplot(x = year.women, y = time.women,
      geom = c("point", "line"),
      xlab = "Year", ylab = "Time",
      main = "100m times for women")

```


#### Compute and plot  proportional decreases in time for each record
```{r}
prop.dec.men = diff(time.men) / time.men[-length(time.men)]

qplot(x = year.men[-length(year.men)], y = prop.dec.men,
      xlab = "Year", ylab = "Proportional decrease",
      main = "100m times for men")
```

```{r}
prop.dec.women = diff(time.women) / time.women[-length(time.women)]
qplot(x = year.women[-length(year.women)], y = prop.dec.women,
      xlab = "Year", ylab = "Proportional decrease",
      main = "100m times for women")
```

 Let's plot the men with the same vertical range as women, to make comparison easier

```{r}
qplot(x = year.men[-length(year.men)], y = prop.dec.men,
      xlab = "Year", ylab = "Proportional decrease",
      ylim = range(prop.dec.women),
      main = "100m times for men")
```




