---
title: "Visualizing Data"
author: "Data Science 101 Team"
output:
  slidy_presentation:
    css: styles.css
    keep_md: yes
  ioslides_presentation: default
---


```{r setup, include=FALSE}
#source('http://datascience101.stanford.edu/profile.R')
knitr::opts_chunk$set(echo = TRUE,cache = TRUE)
library(pacman)
pacman::p_load(Hiiragi2013,tidyverse)
```
## Two packages

- Base R
- ggplot

<p></br></p>
<p></br></p>
<p></br></p>

---------------------------------
The materials of this lecture are taken from the book 'Modern statistics for modern biology' by Susan Holmes and Wolfgang Huber.
---------------------------------

## Base R plotting

- Data from an enzyme-linked immunosorbent assay (ELISA) assay.

```{r}
head(DNase)
plot(DNase$conc,DNase$density)
```

## Base R plotting: a customized version

```{r}
plot(DNase$conc, DNase$density,
  ylab = attr(DNase, "labels")$y,
  xlab = paste(attr(DNase, "labels")$x, attr(DNase, "units")$x),
  pch = 3,
  col = "blue")
```

## Base R plotting: histogram and boxplot

```{r basicplotting3, fig.width=5.6, fig.height=3.5}
hist(DNase$density, breaks=25, main = "")
```

```{r basicplotting4, fig.width=5.6, fig.height=3.5}
boxplot(density ~ Run, data = DNase)
```

## ggplot: Grammar of graphics

- Layering out the graphics
- Plot 1D, 2D, 3-5D data
- Faceting

## An example dataset
```{r loadHiiragi, warning = FALSE,message = FALSE,}
library("Hiiragi2013")
data("x")
dim(exprs(x))
head(pData(x),n=2)
```

## An example dataset - continued
```{r groupSize, warning = FALSE,message = FALSE,}
library("dplyr")
library("ggplot2")
groups = group_by(pData(x), sampleGroup) %>%
  summarise(n = n() , color = unique(sampleColour))
groups
ggplot(data = groups, aes(x = sampleGroup, y = n)) +
  geom_bar(stat = "identity")
```

## An example dataset - a better picture
```{r}
groupColor = setNames(groups$color, groups$sampleGroup)
ggplot(groups, aes(x = sampleGroup, y = n, fill = sampleGroup)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = groupColor, name = "Groups") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Layers
```{r}
dftx = data.frame(t(exprs(x)), pData(x))
ggplot( dftx, aes( x = X1426642_at, y = X1418765_at)) +
  geom_point( shape = 1 ) +
  geom_smooth( method = "loess" )
```

```{r}
ggplot( dftx, aes( x = X1426642_at, y = X1418765_at ))  +
  geom_point( aes( color = sampleColour), shape = 19 ) +
  geom_smooth( method = "loess" ) +
  scale_color_discrete( guide = FALSE )
```

## Visualization of 1D Data

- Barplots
- Boxplots
- Violin plots
- Dot plots and beeswarm plots
- Density plots

## Visualization of 1D Data

A common task in biological data analysis is the comparison between several samples
of univariate measurements. As an example to visulize and compare such samples, we'll use the intensities of a set of four genes Fgf4, Gata4, Gata6 and Sox2. On the array, they are represented by

```{r genes2ps1}
selectedProbes = c( Fgf4 = "1420085_at", Gata4 = "1418863_at",
                   Gata6 = "1425463_at",  Sox2 = "1416967_at")
```

Convert them into a dataframe:

```{r melt}
library("reshape2")
genes = melt(exprs(x)[selectedProbes, ], varnames = c("probe", "sample"))
head(genes)
```

For good measure, we also add a column that provides the gene symbol along with the probe identifiers.

```{r symbol}
genes$gene = names(selectedProbes)[ match(genes$probe, selectedProbes) ]
```

## Barplots and Boxplots

- What information do we get from the graphs?

```{r onedbp1, fig.width = 3, fig.height = 3.75}
ggplot(genes, aes( x = gene, y = value)) +
  stat_summary(fun.y = mean, geom = "bar")
```

```{r onedboxpl, fig.width = 3.75, fig.height = 3.75}
p = ggplot(genes, aes( x = gene, y = value, fill = gene))
p + geom_boxplot()
```


## Violin plots

A variation of the boxplot idea, but with an even more direct representation of the shape of the data distribution, is the violin plot. Here, the shape of the violin gives a rough impression of the distribution density.

```{r onedviolin, fig.width = 3.75, fig.height = 3.75}
p + geom_violin()
```

## Dot plots

When the number of data points is not too large, it is possible to show the data points directly.

```{r oneddot, fig.width = 5, fig.height = 5}
p + geom_dotplot(binaxis = "y", binwidth = 1/6,
       stackdir = "center", stackratio = 0.75,
       aes(color = gene))
```

## Density plots

Choice of size of smoothing window

- small: capture peaks in the dense regions, but may lead to instable (``wiggly'') estimates elsewhere
- big:  sharp peaks, may be smoothed out.

```{r oneddens, fig.width = 4.75, fig.height = 4.75}
ggplot(genes, aes( x = value, color = gene)) + geom_density()
```


## Visualization of 2D data: scatterplots

- Scatterplots are useful for visualizing associations between variables.
- Let us take a look at differential expression between a wildtype and an FGF4-KO sample.

```{r twodsp1, fig.width = 3.75, fig.height = 3.75, dev = "png"}
dfx = as.data.frame(exprs(x))
scp = ggplot(dfx, aes(x = `59 E4.5 (PE)` ,
                      y = `92 E4.5 (FGF4-KO)`))
scp + geom_point()
```


## Visualization of 2D data: scatterplots

- Adjust the transparency

```{r twodsp2, fig.width = 3.75, fig.height = 3.75, dev = "png"}
scp  + geom_point(alpha = 0.1)
```

- 2D density estimate

```{r twodsp3, fig.width = 3.75, fig.height = 3.75}
scp + geom_density2d()
```
What's wrong with this 2D density graph?

## Visualization of 2D data: 2D density estimate

- tweak the bandwidth and binning parameters

```{r twodsp4, fig.width = 3.75, fig.height = 3.75}
scp + geom_density2d(h = 0.5, bins = 60)
```

- with color filling

```{r twodsp5, fig.width = 5.25, fig.height = 3.75, message = FALSE}
library("RColorBrewer")
colorscale = scale_fill_gradientn(
    colors = rev(brewer.pal(9, "YlGnBu")),
    values = c(0, exp(seq(-5, 0, length.out = 100))))

scp + stat_density2d(h = 0.5, bins = 60,
          aes( fill = ..level..), geom = "polygon") +
  colorscale + coord_fixed()
```

## 3-5D data: five aesthetics provided by ggplot

- fill
- color
- shape
- size
- alpha

## 3-5D data: Faceting
We can show multiple plots that result from repeatedly subsetting (or ``slicing'') our data based on one (or more) of the variables, so that we can visualize each part separately. 

```{r facet1, fig.width = 8, fig.height = 2}
library("magrittr")
dftx$lineage %<>% sub("^$", "no", .)
dftx$lineage %<>% factor(levels = c("no", "EPI", "PE", "FGF4-KO"))

ggplot(dftx, aes( x = X1426642_at, y = X1418765_at)) +
  geom_point() + facet_grid( . ~ lineage )
```

## 3-5D data: Faceting

We can specify two faceting variables as well

```{r facet2, fig.width = 8, fig.height = 6}
ggplot( dftx,
  aes( x = X1426642_at, y = X1418765_at)) + geom_point() +
   facet_grid( Embryonic.day ~ lineage )
```

