---
title: "Nonparametric Statistics: a real example about rankings"
author: "Data Science Team"
output: 
       slidy_presentation:
        css: styles.css
        keep_md: true
---

```{r setup, include=FALSE}
#source('http://datascience101.stanford.edu/profile.R')
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction: World University Rankings
This is a great dataset available through [Kaggle.com](https://www.kaggle.com/mylesoneill/world-university-rankings#).

* 3 sources of rankings are included: Times rankings, center for world university rankings, and Shanghai rankings.

* We will spend sometime on data cleaning and visualization.

* Use PCA to see which covariates drive the rankings the most. Indeed, we could use principal component regression, using the PCA scores as covariates in the model.

* Barro-Lee Dataset on Educational Attainment.


## Setup
``` {r libraries, warning = FALSE, message = FALSE}
library(maps)
library(magrittr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(gridExtra)
```

## Let us look at each of the dataset 

- Times Higher Education World University Ranking [(source)](https://www.timeshighereducation.com/world-university-rankings)

- timesData.csv
      + world_rank - world rank for the university. Contains rank ranges and equal ranks (eg. =94 and 201-250).
      + university_name - name of university.
      + country - country of each university.
      + teaching - university score for teaching (the learning environment).
      + international - university score international outlook (staff, students, research).
      + research - university score for research (volume, income and reputation).
      + citations - university score for citations (research influence).
      + income - university score for industry income (knowledge transfer).
      + total_score - total score for university, used to determine rank.
      + num_students - number of students at the university.
      + student_staff_ratio - Number of students divided by number of staff.
      + international_students - Percentage of students who are international.
      + female_male_ratio - Female student to Male student ratio.
      + year - year of the ranking (2011 to 2016 included).
      
## Load the data and rename some categories      
```{r loadTimes, warning = FALSE}
timesdf <- read.csv("./data/timesData.csv", na.strings = '-', stringsAsFactors = FALSE )
timesdf %<>% 
  tbl_df %>%
  rename( institution = university_name ) %>% 
  mutate(
    institution = gsub(" *(-|–) *","-",institution),
    institution = gsub("Leibniz University of Hanover","Leibniz University of Hannover",institution),
    institution = gsub("William & Mary","College of William and Mary",institution),
    institution = gsub("São Paulo State University (UNESP)","UNESP, São Paulo State University",institution),
    institution = gsub("University of Paris North-Paris 13","University of Paris Nord (Paris 13)",institution),
    institution = gsub("Louisiana State University","Louisiana State University-Baton Rouge",institution),
    female_male_ratio = as.numeric((gsub('^([0-9]+).+','\\1',female_male_ratio)))/100,
    international_students = as.numeric(gsub('%','',international_students))/100,
    num_students = as.numeric(gsub(',','',num_students)),
    country = gsub('Republic of Ireland','Ireland',country),
    country = gsub('Russian Federation','Russia',country),
    country = gsub("Unisted States of America",'United States of America',country),
    country = gsub("Unted Kingdom","United Kingdom",country)
  ) %>%
  rename(femaleProp = female_male_ratio)
```

## Brief summary
```{r}
summary(timesdf)
```

## Center for World University Rankings [(source)](http://cwur.org/)

*  cwurData.csv
      + world_rank - world rank for university.
      + university_name - name of university.
      + country - country of each university.
      + national_rank - rank of university within its country.
      + quality_of_education - rank for quality of education.
      + alumni_employment - rank for alumni employment.
      + quality_of_faculty - rank for quality of faculty.
      + publications - rank for publications.
      + influence - rank for influence.
      + citations - rank for citations.
      + broad_impact - rank for broad impact (only available for 2014 and 2015)
      + patents - rank for patents.
      + score - total score, used for determining world rank.
      + year - year of ranking (2012 to 2015).
   
##   Load the data and rename some categories  

```{r loadCWUR, warning = FALSE}
cwurdf <- read.csv("./data/cwurData.csv", stringsAsFactors = FALSE )
cwurdf %<>%
  tbl_df %>%
  mutate(
    institution = gsub(" *(-|–) *","-",institution),
    institution = gsub(" catholique"," Catholique",institution),
    institution = gsub("University of Hanover","Leibniz University of Hannover",institution),
    institution = gsub("Western University (The University of Western Ontario)","Western University of Ontario",institution),
    institution = gsub("^Purdue University.+","Purdue University",institution),
    institution = gsub("Royal Veterinary College, University of London","Royal Veterinary College",institution),
    institution = gsub("University of Pittsburgh-Pittsburgh Campus","University of Pittsburgh",institution),
    institution = gsub("Texas A&M University, College Station","Texas A&M University",institution),
    institution = gsub("University of Paris-Sud","University of Paris Sud (Paris 11)",institution),
    institution = gsub("University of Paris 13","University of Paris Nord (Paris 13)",institution),
    institution = gsub("London School of Economics","London School of Economics and Political Science",institution),
    country = gsub('Slovak Republic','Slovakia',country),
    country = gsub('^USA$','United States of America',country),
    country = gsub('^UK$','United Kingdom',country)
  )
```

## Brief summary
```{r}
summary(cwurdf)
```

## Academic Ranking of World Universities 
[(source)](http://www.shanghairanking.com/) (aka the Shanghai Rankings)  

*  shanghaiData.csv
      + world_rank - world rank for university. Contains rank ranges and equal ranks (eg. 101-152).
      + university_name - name of university.
      + national_rank - rank of university within its country.
      + total_score - total score, used to determine rank.
      + alumni - Alumni Score, based on the number of alumni of an institution winning nobel prizes and fields medals.
      + award - Award Score, based on the number of staff of an institution winning Nobel Prizes in Physics, Chemistry, Medicine, and Economics and Fields Medals in Mathematics.
      + hici - HiCi Score, based on the number of Highly Cited Researchers selected by Thomson Reuters.
      + ns - N&S Score, based on the number of papers published in Nature and Science.
      + pub - PUB Score, based on total number of papers indexed in the Science Citation Index-Expanded and Social Science Citation Index.
      + pcp - PCP Score, the weighted scores of the above five indicators divided by the number of full time academic staff.
      + year - year of ranking (2005 to 2015).

##  Load the data and clean.        
```{r loadShanghai, warning = FALSE}
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
# Determine the country associated with each university in the Shanghai dataset.
schoolCountry <- read.csv("./data/school_and_country_table.csv", stringsAsFactors = FALSE) %>% tbl_df
schoolCountry %<>% 
  mutate(
    school_name = gsub(" *(-|–) *","-",school_name),
    school_name = gsub("Leibniz University of Hanover","Leibniz University of Hannover",school_name),
    country = gsub('Republic of Ireland','Ireland',country),
    country = gsub('Russian Federation','Russia',country),
    country = gsub('Slovak Republic','Slovakia',country)
  ) %>%
  bind_rows(timesdf %>% dplyr::select(institution,country) %>% rename(school_name = institution)) %>%
  bind_rows(cwurdf  %>% dplyr::select(institution,country) %>% rename(school_name = institution)) %>%
  distinct

shanghaidf <- read.csv("./data/shanghaiData.csv", stringsAsFactors = FALSE ) %>% tbl_df
shanghaidf %<>% filter(university_name != "")
```

## Reconcile university names with those of timesdf and cwurdf

```{r}
substPatterns <- matrix(
  c(
    "Aix Marseille","Aix-Marseille",
    "Hygiene Tropical","Hygiene & Tropical",
    "A *& *M","A&M",
    "^The ","",
    '^University of Queensland','The University of Queensland',
    '^Open University','The Open University',
    "Vrije (University|Universiteit) Brussel.*","Vrije Universiteit Brussel",
    "(University Libre Bruxelles)|(Université libre de Bruxelles.*)","Université Libre de Bruxelles",
    "Queen Mary, (U\\.|University) of London","Queen Mary University of London",
    " *- *","-",
    "Curtin University of Technology","Curtin University",
    "University of Hannover","Leibniz University of Hannover",
    "College of +William","College of William",
    "Erasmus University","Erasmus University Rotterdam",
    "Western University (The University of Western Ontario)","University of Western Ontario",
    "^Western University$","University of Western Ontario",
    "Imperial College of Science, Technology and Medicine","Imperial College London",
    "Purdue University-West Lafayette","Purdue University",
    "^UNESP$","UNESP, São Paulo State University",
    "MINES P","Mines P",
    "(Korea Advanced Institute of Science and Technology)","\\1 (KAIST)",
    "Ecole Polytechnique","École Polytechnique",
    "University of New England","University of New England (Australia)",
    "University of California-Berkeley","University of California, Berkeley",
    "University of Pittsburgh.*","University of Pittsburgh",
    "Texas A&M University-College Station","Texas A&M University",
    "University of Paris-Sud (Paris 11)","University of Paris Sud (Paris 11)",
    "University of Roma-La Sapienza","Sapienza University of Rome",
    "University of Santiago Compostela","University of Santiago de Compostela",
    "Massachusetts Institute of Technology.+","Massachusetts Institute of Technology",
    "London School of Hygiene and Tropical Medicine","London School of Hygiene & Tropical Medicine"
  )
  ,ncol=2,
  byrow=TRUE,
  dimnames = list(NULL,c('from','to'))
)

schools <- shanghaidf$university_name
for(i in seq_len(nrow(substPatterns))){
  schools <- gsub(substPatterns[i,1],substPatterns[i,2],schools)
}
shanghaidf$university_name <- schools

```
## Fix school and country names

Load the `world.cities` dataset from package `maps`.

```{r loadcities}
data("world.cities")
wc <- world.cities %>% 
  tbl_df %>% 
  arrange(desc(pop)) %>%
  filter( !duplicated(name) ) %>%
  dplyr::select(name,country.etc) %>% 
  mutate(country.etc = gsub('^UK$','United Kingdom',gsub('^USA$','United States of America',country.etc)) ) %>%
  rename(
    city    = name,
    country = country.etc
  ) %>%
  filter( !(city == 'Roma' & country == 'Australia')) %>%
  filter( !(city == 'Santiago' & country == 'Philippines')) %>%
  filter( !(city == 'Halle' & country == 'Germany')) %>%
  bind_rows(
    data_frame(
      city = c("Goettingen","Muenster","Duesseldorf","Roma","Yang Ming","Mount Sinai","Norwegian","King Fahd","ParisTech"),
      country = c("Germany","Germany","Germany","Italy","China","United States of America","Norway","Saudi Arabia","France")
    )
  )
countries <- wc$country %>% unique %>% sort
```

## Then assign countries to Shanghai dataset.
```{r asgncntry}
# First find the exact matches for university name in the `schoolCountry` dataset.
set1 <- shanghaidf$university_name %>% unique %>% sort
a <- schoolCountry %>% filter( school_name %in% set1 )
set1 %<>% setdiff(a$school_name)
# Next, get the schools that are obviously American because they have a US state in their name.
a %<>% bind_rows(
  data_frame(
    school_name = state.name %>% paste(collapse="|") %>% grep(set1) %>% { set1[ . ] },
    country = 'United States of America' 
  ) 
)
set1 %<>% setdiff(a$school_name)
# Then, get the schools that have countries in their names
countries <-
a %<>% bind_rows(
  data_frame(
    school_name = countries %>% sapply(function(cntry) cntry %>% grep(set1)) %>% unlist %>% { set1[ . ] },
    country    =  countries %>% sapply(function(cntry) cntry %>% grep(set1) %>% { rep( cntry, length( . ) ) } ) %>% unlist
  ) 
)
set1 %<>% setdiff(a$school_name)
```

## Then, get the schools that have a world city in their names
```{r}
ischool    <- wc$city %>% sapply(function(city) city %>% { paste0("(^|[- (])", . ,"($|[- )])") } %>% grep(set1))
ischool    <- ischool[sapply(ischool,length)>0]
cntryNames <- setNames(wc$country,wc$city)[names(ischool)]
a %<>% bind_rows(
  data_frame(
    school_name = set1[ ischool %>% unlist ],
    country     = rep(cntryNames,times=sapply(ischool,length))
  ) 
)
set1 %<>% setdiff(a$school_name)
# Now strip down the unassigned schools to just the place/name and see if a school with an assigned country
# also has the same place/name and therefore belongs to the same country.

shanghaidf2 <- shanghaidf %>% 
  full_join(a, by = c('university_name' = 'school_name')) %>% 
  dplyr::select(1:2,country,everything()) %>%
  rename( institution = university_name ) 
```


## Combine Rankings

Combine the rankings into a single data.frame

```{r combine, warning = FALSE, fig.width = 11, fig.height = 9}
countries <- timesdf$country %>% 
  union( cwurdf$country ) %>% 
  union( shanghaidf2$country ) %>% 
  sort %T>% print
```

## Combine Rankings

```{r}
combodf <- timesdf %>% 
  full_join( shanghaidf2, by = c('institution','country','year') ) %>% 
  full_join(     cwurdf,  by = c('institution','country','year') ) %>%
  mutate(
    institution = factor(institution),
    country     = factor(country)
  ) %>%
  rename(
      rank_times        = world_rank.x,
      rank_shanghai     = world_rank.y,
      rank_cwur         = world_rank,
      citations_times   = citations.x,
      citations_cwur    = citations.y,
      score_times       = total_score.x,
      score_shanghai    = total_score.y,
      natlrank_shanghai = national_rank.x,
      natlrank_cwur     = national_rank.y
  ) %>%
  dplyr::select(institution,country,year,starts_with('rank'),matches('rank'),everything()) %T>% { summary(.) %>% print }
```

## Show the counts of institutions captured in each source-year combo.

```{r cntinst}
countRanks <- function(df){
  sapply(
    2012:2015 %>% 
      setNames(2012:2015),
    function(yr) { 
      df %>% 
        filter(year==yr) %$% 
        { 
          c(
            times=length(unique(rank_times)),
            cwur=length(unique(rank_cwur)),
            shanghai=length(unique(rank_shanghai))
          ) 
        }
    }
  )
}

combodf    %>% countRanks %T>% print %>% sum
```
## If we only want to see the top-100 rankings:
```{r}
combo100df <- combodf %>% filter((rank_times %in% as.character(1:100)) | (rank_cwur < 101) | 
(rank_shanghai %in% as.character(1:100)))
```

```{r}
combo100df %>% countRanks %T>% print %>% sum
```

## Sort according to rankings
```{r}
timesdf2 <- timesdf %>% mutate(world_rank = sprintf('%8s',world_rank))
tmp.times <- lapply(unique(timesdf2$year) %>% sort, 
              function(i) {
                tmp<- timesdf2 %>% filter(year==i); 
                ry <- rank(tmp$world_rank); 
                tmp %<>% mutate(
                  rank.times=ry,
                  qrank.times=1-ecdf(ry)(ry))
              }
) %>% { do.call( rbind, . ) }
tmp.cwur <- lapply(unique(cwurdf$year) %>% sort, 
                    function(i) {
                      tmp<- cwurdf %>% filter(year==i); 
                      ry <- rank(tmp$world_rank); 
                      tmp %<>% mutate(
                        rank.cwur=ry,
                        qrank.cwur=1-ecdf(ry)(ry))
                    }
) %>% { do.call( rbind, . ) }
shanghaidf2 %<>% mutate(world_rank = sprintf('%8s',world_rank))
tmp.shanghai <- lapply(unique(shanghaidf2$year) %>% sort, 
                    function(i) {
                      tmp<- shanghaidf2 %>% filter(year==i); 
                      ry <- rank(tmp$world_rank); 
                      tmp %<>% mutate(
                        rank.shanghai=ry,
                        qrank.shanghai=1-ecdf(ry)(ry))
                    }
) %>% { do.call( rbind, . ) }
```

## Visualize Combined Data

Just a quick check on what we've got. Rankings color-coded by country.

```{r pltcombo, fig.width=12, fig.height=8}
iyr <- 2015
gplt.times <- tmp.times %>% 
  dplyr::select(1:3,year,contains('rank')) %>% 
  filter(year == iyr,qrank.times>0.9) %>% 
  mutate(institution = factor(institution,levels=unique(institution)[order(qrank.times,decreasing=TRUE)])) %>% 
  ggplot(aes(x=institution,y=qrank.times,fill=country)) + 
  geom_bar(stat='identity') + 
  theme(axis.text.x = element_text(angle = 90,hjust=1.0)) + labs(title = sprintf('Times World Rankings %d',iyr))

gplt.cwur <- tmp.cwur %>% 
  dplyr::select(1:3,year,contains('rank')) %>% 
  filter(year == iyr,qrank.cwur>0.9) %>% 
  mutate(institution = factor(institution,levels=unique(institution)[order(qrank.cwur,decreasing=TRUE)])) %>% 
  ggplot(aes(x=institution,y=qrank.cwur,fill=country)) + 
  geom_bar(stat='identity') + 
  theme(axis.text.x = element_text(angle = 90,hjust=1.0)) + labs(title = sprintf('CWUR World Rankings %d',iyr))

gplt.shanghai <- tmp.shanghai %>% 
  dplyr::select(1:3,year,contains('rank')) %>% 
  filter(year == iyr,qrank.shanghai>0.9) %>% 
  mutate(institution = factor(institution,levels=unique(institution)[order(qrank.shanghai,decreasing=TRUE)])) %>% 
  ggplot(aes(x=institution,y=qrank.shanghai,fill=country)) + 
  geom_bar(stat='identity') + 
  theme(axis.text.x = element_text(angle = 90,hjust=1.0)) + labs(title = sprintf('Shanghai World Rankings %d',iyr))

marrangeGrob(list(times = gplt.times, cwur = gplt.cwur, shanghai = gplt.shanghai), nrow = 1, ncol = 3)
```


## Bayesian Plackett-Luce Modeling to infer institution strengths from rankings

- Starting with the combo top-100 rankings, combo100df, convert into an ordering matrix with the appropriate covariate matrices for X, country, source, year.

- Here, we'll use the inverse-logistic (effectively, log-odds) of the quantiles of each 
variable from the 3 ranking sources (Times, CWUR, Shanghai). This is convenient since some of the variables are scores and others are rankings.  
```{r conversion1}
a <- combo100df %>% 
  dplyr::select(-contains('score')) %>% 
  mutate(schoolID  = as.integer(factor(institution)),
         countryID = as.integer(factor(country)))
```
## Getting the covariates, X
```{r}
medianMLT <- function(x) median(x,na.rm=TRUE)
instX <- a %>% 
  group_by(institution) %>% 
  summarize_each(funs(medianMLT),teaching:countryID) %>%
  inner_join(a %>% dplyr::select(country,countryID) %>% distinct,by='countryID')
```
Variable `femaleProp` has `NA` for quite a few institutions, so we'll drop it.
In order to retain institutions like Columbia U. and Purdue U., we'll drop the `income` column.

```{r dropit}
instX %<>% dplyr::select(-femaleProp,-income)

# *************************************************
# ***Drop any row with 6 or more missing values.***
# ***THIS DROPS ANY UNIVERSITY THAT DIDN'T SHOW ***
# *** UP ON AT LEAST TWO OF THE RANKING LISTS   ***
# *************************************************
instX %<>% filter(apply(.,1,function(x) length(which(is.na(x)))<6))
# We'll drop these schools too.
instX %>% filter(!complete.cases(.)) %>% print
instX %<>% filter(complete.cases(.)) %>%
  mutate(
    country     = factor(as.character(country)),
    countryID   = as.integer(country),
    institution = factor(as.character(institution)),
    schoolID    = as.integer(institution)
  )
a     %<>% filter(institution %in% instX$institution) %>% 
  mutate(
    country     = factor(as.character(country)),
    countryID   = as.integer(country),
    institution = factor(as.character(institution)),
    schoolID    = as.integer(institution)
  )
```

This leaves `r nrow(instX)` institutions.


## Continue with the column conversions...
```{r conversion2}
# Convert all columns into log-odds of empirical CDF.
# The rankings variables are negated so that less negative values are better (or could have reversed the ranking values if 
# wanted to keep everything positive).
negate <- function(x) -x
makeBayesFactors <- function(x) qlogis(rank(x)/(length(x)+1))
instX %<>% 
  mutate_each(funs(negate),one_of('quality_of_education','alumni_employment','quality_of_faculty','publications','influence','citations_cwur','patents')) %>%
  mutate_each(funs(makeBayesFactors),-institution,-country,-schoolID,-countryID) %>% 
  dplyr::select(institution,country,schoolID,countryID,everything())

# So this is now our design matrix of covariate values. 
# One row for each of the 113 institutions, and one column for each
# of the 15 covariates we've decided to keep.
X <- instX %>% 
  arrange(schoolID) %>% 
  dplyr::select(-institution,-schoolID,-country,-countryID) %>% 
  dplyr::select(-citations_cwur) %>% # Also drop citations_cwur because have citations_times
  as.matrix
rownames(X) <- instX %>% arrange(schoolID) %$% { institution %>% as.character }
```

## Visualization of the Covariates

We use principal components analysis (PCA) to display the level of correlation amongst the covariates and to hint at which might drive the rankings the most.

Considering that Harvard, Stanford and MIT are near the top of all of the rankings, we expect that the biplot indicates `influence`, `research`, `hici`, `quality_of_education`, `quality_of_faculty`, `alumni` and `award` to have high potential as drivers of the rankings.

## Visualization of the Covariates

```{r pcaX, fig.width = 11, fig.height=8}
xpca <- princomp(X, cor = FALSE)
plot(xpca)
```

## Visualization of the Covariates
```{r}
biplot(xpca)
```

## PCA results
```{r}
print(xpca)
```

## PCA results with names
```{r}
scl2char <- function(x) {xmax<- x[which.max(abs(x))]; if(xmax<0) fctr<- -1 else fctr <- 1; ifelse(abs(x)>0.25,sprintf("%4.2f",fctr*x),"")}
xpcachar <- xpca$loadings %>% matrix(nrow=nrow(.),dimnames=dimnames(.)) %>%
  as.data.frame %>% 
  tbl_df %>% 
  mutate_each(funs(scl2char)) %>% 
  mutate(X=rownames(xpca$loadings)) %>% 
  dplyr::select(X,everything()) %T>% print(width=Inf)

X <- xpca$scores %>% apply(2,function(x) {xmax<- x[which.max(abs(x))]; if(xmax<0) fctr<- -1 else fctr <- 1; fctr*x })

```
## Another related interesting dataset about educational attainment

* UNESCO Institute for Statistics and the Barro-Lee Dataset on Educational Attainment
[(SOURCE)](http://datatopics.worldbank.org/Education/wDataQuery/QProjections.aspx)

*  educational_attainment_supplementary_data.csv
      + Barro-Lee Dataset (2014) - Average years of schooling; Educational attainment among age groups and genders; Population distributions. 144 countries. 1985-2010 in 5 year intervals.
      + UNESCO Institute for Statistics (UIS) - Mean years of schooling; Percentage of population (age 25+) by educational attainment. 152 countries. 1985-2015.
      
## Here, we will focus on the Barro-Lee Dataset
      
```{r loadAttain}
attaindf <- read.csv("./data/educational_attainment_supplementary_data.csv", stringsAsFactors = FALSE)
attaindf %<>% 
  tbl_df %>%
  filter( series_name != "") %>%
  mutate(
    series_name = factor(series_name),
    country_name = factor(country_name)
  )
```

## Stack all the columns across the years into a `Year` and `Value` column (using `tidyr::gather()`).

```{r}
attaindf %<>%
  gather( key = Year, value = Value, -country_name, -series_name ) %>%
  mutate(
    Year      = as.numeric( gsub('X','',Year))
  ) %>%
  filter(!is.na(Value)) %T>% { summary(.) %>% print }
```



## Barro-Lee Data

There's quite a bit of information to work through in this dataset.  We began by breaking off chunks and transforming and plotting them to get a feel for what the
data were saying. Let us first look at the Barro-Lee average years of schooling and population data.
```{r bldf}
bldf <- attaindf %>% filter(grepl('^Barro',series_name))
```

## Average years of schooling
```{r}
bldfay <- bldf %>% filter(grepl('Average years',series_name)) %>%
  mutate(
    source = 'Barro-Lee',
    type = "Average years of schooling",
    schoolLevel = factor(gsub('.+ ([^ ]+) schooling.+','\\1',series_name),levels=c('primary','secondary','tertiary','total')),
    ageCohort   = factor(gsub('.+, age ([^,]+),.+','\\1',series_name)),
    gender      = factor(gsub('.+, ([a-zA-Z]+)$','\\1',series_name))
  ) %>%
  dplyr::select(-series_name)
```

## Proportions finished schooling
```{r}
bldfpf <- bldf %>% filter(grepl('Percentage of',series_name)) %>%
  mutate(
    Value = Value/100,
    source = 'Barro-Lee',
    type = "Proportion with schooling",
    schoolLevel = factor(gsub('.+\\with (.+)$','\\1',series_name),
                         levels=c(
                           'no education',
                           'primary schooling. Completed Primary',
                           'primary schooling. Total (Incomplete and Completed Primary)',
                           'secondary schooling. Completed Secondary',
                           'secondary schooling. Total (Incomplete and Completed Secondary)',
                           'tertiary schooling. Completed Tertiary',
                           'tertiary schooling. Total (Incomplete and Completed Tertiary)'
                         )
    ),
    ageCohort   = factor(gsub('.+ age ([^ ]+) .+','\\1',series_name)),
    gender      = factor(ifelse(gsub('.+Percentage of ([a-zA-Z]+) .+','\\1',series_name)=='population','total','female'))
  ) %>%
  dplyr::select(-series_name)
```

## Population
```{r}
bldfpop <- bldf %>% filter(grepl('Population',series_name)) %>%
  mutate(
    source = 'Barro-Lee',
    type = "Population in thousands",
    ageCohort   = factor(gsub('.+, age ([^,]+),.+','\\1',series_name)),
    gender      = factor(gsub('.+, ([a-zA-Z]+)$','\\1',series_name)),
    population  = Value*1000
  ) %>%
  dplyr::select(-series_name,-Value)

bldfay %<>% inner_join(bldfpop %>% dplyr::select(-source,-type), by = c("country_name","Year","ageCohort","gender"))
bldfpf %<>% inner_join(bldfpop %>% dplyr::select(-source,-type), by = c("country_name","Year","ageCohort","gender"))
```

## Plot out the data to get a feel for its structure and message.
```{r usplots1, fig.width = 10, fig.height = 8}
country <- "United States"
bldfay %>% 
  filter(
    country_name == country,
    !grepl('15\\+|25\\+',ageCohort)
  ) %>% 
  ggplot(aes(x = Year, y = Value, color = gender)) + 
     geom_line() + labs(y = 'Average Years of Schooling', title = country) +
     theme(axis.text.x = element_text(angle = 90)) +
     facet_wrap(schoolLevel~ageCohort, nrow = 4)
```

## Plot out the data to get a feel for its structure and message.

```{r}
country <- "United States"
bldfay %>%
  filter(
    country_name == country,
    !grepl('15\\+|25\\+',ageCohort)
  ) %>%
  ggplot(aes(x = Year, y = Value, color = ageCohort)) +
  geom_line() + labs(y = 'Average Years of Schooling', title = country) +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(gender ~ schoolLevel, nrow = 2)
```

